Values Encoded in Machine Learning Research

### 2025-02-13 latex/paper.tex
repository url
add abstract

% Abstract Section
\begin{abstract}
% \textbf{Keywords:} Machine Learning, Ethics, Values, Society, Impact
\end{abstract}

% Sections
\section{Introduction}
Machine learning (ML) research has become a driving force behind technological advancements across diverse domains, including healthcare \cite{esteva2019guide}, finance \cite{berg2022fintech}, and autonomous systems \cite{hawkins2021guidance}. 
However, despite its transformative potential, the values embedded in ML research remain underexamined, 
raising critical questions about whom the field serves and whose needs it prioritizes. The previous study by Birhane et al. \cite{valuesInML2021} has demonstrated that ML research overwhelmingly prioritizes 
performance, generalization, and novelty while often neglecting ethical considerations, fairness, and broader societal impacts.

@@

\section{Methodology}
\subsection{Data collection}
We curated a dataset of the 100 most influential papers from NeurIPS 2023 and ICML 2023 using rankings provided by Paper Digest 
(\url{https://www.paperdigest.org/topic/?topic=nips&year=2023} and \url{https://www.paperdigest.org/topic/?topic=icml&year=2023}). 
These rankings were based on an impact score, which reflects a combination of paper citations, patent citations, etc. It is a score in 1.0-10.0, 
with a higher value indicating a broader impact. We then downloaded the respective papers from the conference website directly 
(\url{https://proceedings.neurips.cc/paper_files/paper/2023} and \url{https://icml.cc/Downloads/2023}). 
The selected papers represent the cutting edge of machine 
learning research and provide a representative sample of current trends and values in the field. We did not use Semantic Scholar as in the Birhane et 
al. (2023) paper because Semantic Scholar did not provide API keys due to high demand. We could also not use Google Scholar to access the citation 
counts because the IP address gets blocked after too many access tries. A list of the PDFs of the 100 papers of both conferences can be found in the 
GitHub repository.

@@

\subsection{Quantitative Summary}
We measure the prevelence of each keyword in the 100 papers from the ICML 2023 conference and the 100 papers from the NeurIPS 2023 conference. First we investigate the combined distribution of the keywords in both conferences, second we look at both conferences seperately,
and lastly we compare the results from the 2023 papers with the results from the 2008/09 and 2018/19 papers from Birhane et al. \cite{valuesInML2021}. 
In line with the Birhane et al. paper we highlight the values that fall under `User rights'. These are `Interpretable (To Users)', `Deferral To Humans', `Privacy', `User influence', `Not Socially Biased' and `Fairness'. 
These values are highlighted in all plots in orange. The second group of values highlighted are values belonging to `Ethical Principles', These keywords are `Beneficence', `Non-Maleficence', `Respect for Law and Public Interest', 
`Respect for Persons', `Autonomy (Power to Decide)', `Explicability' and `Justice'.
\subsubsection{ICML and Neurips 2023}
In Figure \ref{fig:percentage_both_conferences} we show the prevelance of values in all 200 papers combined from ICML and Neurips 2023. The top values are performance (99\%), accuracy (85\%), 
Understanding (for researchers) (82.5\%), Efficiency (81\%) and Fast (80\%). Among the values related to user rights and stated in 
ethical principles the most common ones are `Explicability' (22.5\%) and Privacy (21.5\%).

@@

>>>>

% Abstract Section
\begin{abstract}
% \textbf{Keywords:} Machine Learning, Ethics, Values, Society, Impact
Machine learning (ML) research drives technological progress across various domains, yet the values encoded in this research remain underexplored. Prior studies suggest that ML research overwhelmingly prioritizes 
technical performance over ethical considerations, fairness, and broader societal impacts. Building upon the work of Birhane et al.\cite{valuesInML2021}, we analyze the value commitments in contemporary ML research by examining 
the most influential papers from NeurIPS 2023 and ICML 2023. Using a keyword-based and qualitative analysis, we assess the prevalence of ethical and performance-driven values in these papers. Our findings reveal that while 
performance, efficiency, and accuracy continue to dominate, there is an increasing—but still limited—acknowledgment of values related to user rights and ethical principles, particularly privacy and explicability. However, 
fairness, transparency, and social responsibility remain secondary concerns. This study highlights the need for a more deliberate integration of ethical considerations into ML research and calls for alternative evaluation metrics 
that balance technical advancements with social responsibility. By providing empirical insights into the values upheld in contemporary ML research, we contribute to the ongoing discourse on the socio-political dimensions of 
AI development and encourage a reorientation towards more equitable and inclusive technological progress.
\end{abstract}

% Sections
\section{Introduction}
Machine learning (ML) research has become a driving force behind technological advancements across diverse domains, including healthcare \cite{esteva2019guide}, finance \cite{berg2022fintech}, and autonomous systems \cite{hawkins2021guidance}. 
However, despite its transformative potential, the values embedded in ML research remain underexamined, 
raising critical questions about whom the field serves and whose needs it prioritizes. Therefore it is cirital to analyse the values encoded in ML research. Brey states that technology is not neutral but has embedded values \cite{brey2010values}. 
The previous study by Birhane et al. \cite{valuesInML2021} has demonstrated that ML research overwhelmingly prioritizes 
performance, generalization, and novelty while often neglecting ethical considerations, fairness, and broader societal impacts.

@@

\section{Methodology}
\subsection{Data collection}
We curated a dataset of the 100 most influential papers from NeurIPS 2023 and ICML 2023 using rankings provided by Paper Digest 
(\url{https://www.paperdigest.org/topic/?topic=nips&year=2023} and \url{https://www.paperdigest.org/topic/?topic=icml&year=2023}). 
These rankings were based on an impact score, which reflects a combination of paper citations, patent citations, etc. It is a score in 1.0-10.0, 
with a higher value indicating a broader impact. We then downloaded the respective papers from the conference website directly 
(\url{https://proceedings.neurips.cc/paper_files/paper/2023} and \url{https://icml.cc/Downloads/2023}). 
The selected papers represent the cutting edge of machine 
learning research and provide a representative sample of current trends and values in the field. We did not use Semantic Scholar as in the Birhane et 
al. (2023) paper because Semantic Scholar did not provide API keys due to high demand. We could also not use Google Scholar to access the citation 
counts because the IP address gets blocked after too many access tries. A list of the PDFs of the 100 papers of both conferences can be found in the 
GitHub repository: \url{https://github.com/IsabelKurth/values-in-ml-research}.

@@

\section{Results}
\subsection{Quantitative Summary}
We measure the prevelence of each keyword in the 100 papers from the ICML 2023 conference and the 100 papers from the NeurIPS 2023 conference. First we investigate the combined distribution of the keywords in both conferences, second we look at both conferences seperately,
and lastly we compare the results from the 2023 papers with the results from the 2008/09 and 2018/19 papers from Birhane et al. \cite{valuesInML2021}. 
In line with the Birhane et al. paper we highlight the values that fall under `User rights'. These are `Interpretable (To Users)', `Deferral To Humans', `Privacy', `User influence', `Not Socially Biased' and `Fairness'. 
These values are highlighted in all plots in orange. The second group of values highlighted are values belonging to `Ethical Principles', These keywords are `Beneficence', `Non-Maleficence', `Respect for Law and Public Interest', 
`Respect for Persons', `Autonomy (Power to Decide)', `Explicability' and `Justice'.
\subsubsection{ICML and Neurips 2023}
In Figure \ref{fig:percentage_both_conferences} we show the prevelance of values in all 200 papers combined from ICML and Neurips 2023. The top values are performance (99\%), accuracy (85\%), 
Understanding (for researchers) (82.5\%), Efficiency (81\%) and Fast (80\%). Among the values related to user rights and stated in 
ethical principles the most common ones are `Explicability' (22.5\%) and Privacy (21.5\%).

@@

We choose the papers we include in our analysis based on the impact score provided by Paper Digest. This score is based on a combination of paper citations, patent citations, etc. What needs to be mentioned here is that
citations always have a time delay and it is hard to determine just shortly after a conference witch papers will have the greatest influence in the future. By including 100 papers per conference, we aim to 
capture the most influencial once but it is possible that some papers that will have a great impact in the future are not included in our analysis. Also citations only might not be a sufficient criteria for the influence of a paper.
